Flute and kink instabilities in a dynamically twisted flux tube with anisotropic plasma viscosity

### 2022-01-17 paper.tex
Text edits for main Q1,2,3

%% --- user packages ---
\usepackage{csquotes}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{grffile}
\usepackage{overpic}
\usepackage{float}
\usepackage{enumitem}
\usepackage{amsfonts}
\usepackage{easy-todo}
\usepackage{listings}
%\usepackage[citecolor=blue]{hyperref}
\usepackage{lineno}

@@

\newcommand{\rev}[1]{{\color{red} #1}}

@@

\begin{abstract}
Magnetic flux tubes such as those in the solar corona are subject to a number
of instabilities. Important among them is the kink instability which plays
a central part in the nanoflare theory of coronal heating, and for this reason
in numerical simulations it is usually induced by tightly-controlled
perturbations and studied in isolation. In contrast, we find that when
disturbances are introduced in our magnetohydrodynamic flux tube simulations by
dynamic twisting of the flow at the boundaries fluting modes of instability are
readily excited. We also find that the flute instability, which has been
theorised but rarely observed in the coronal context,  is strongly enhanced
when plasma viscosity is assumed anisotropic. We proceed to investigate the
co-existence and competition between flute and kink instabilities for a range
of values of the resistivity and of the parameters of the anisotropic and
isotropic models of viscosity. We conclude that while the flute instability
cannot prevent the kink from ultimately dominating, it can significantly delay
its development especially at strong viscous anisotropy induced by intense
magnetic fields.
\end{abstract}

@@

The helical kink instability is a form of ideal magnetohydrodynamic (MHD)
instability which occurs in highly twisted magnetic flux tubes such as those
making up much of the solar corona~\citep{realeCoronalLoopsObservations2014}
and has been well studied in the coronal
context~\citep{hoodKinkInstabilitySolar1979,hoodCoronalHeatingMagnetic2009,browningSolarCoronalHeating2003b,barefordShockHeatingNumerical2015,quinnEffectAnisotropicViscosity2020}.
Given its energetic nonlinear development, it is considered a potential
mechanism for heating the solar corona through the theory of
nanoflares~\citep{klimchukSolvingCoronalHeating2006,browningMechanismsSolarCoronal1991}
and a key mechanism in the production of solar
flares~\citep{hoodKinkInstabilitySolar1979}. \rev{In our previous work
we have} investigated a twisted magnetic flux tube already linearly unstable to the
helical kink instability, focussing specifically on the effect of anisotropic
viscosity on the nonlinear
dynamics~\citep{quinnEffectAnisotropicViscosity2020}.
\rev{There and in} most other investigations of
the kink instability e.g.~\rev{\citep{hoodCoronalHeatingMagnetic2009}},
a perturbation is applied to an already significantly twisted flux tube. An
alternative way to excite the instability (and the way employed here) is to
start with an initially straight field and apply twisting motions at the
boundaries to form a twisted flux tube which eventually becomes unstable. This
kind of dynamic excitation of the kink instability is useful in that it
represents more closely the actual evolution of magnetic flux tubes and the
associated instabilities in the solar corona. In our simulations, the dynamic
twisting of the flux tube reveals an additional instability, the flute
instability, which has been theorised, for example
in~\rev{\citep{priestMagnetohydrodynamicsSun2013}}. While oscillations resembling
flute perturbations have been found in simulations of coronal
loops~\citep{terradasEffectMagneticTwist2018}, to our knowledge, this is the
first time the flute instability has been investigated computationally in
a coronal context.

@@

In other solar contexts, interchange instabilities can be found in the form of
ballooning modes in arcades~\citep{hoodBallooningInstabilitiesSolar1986}, as
the instability which forms tubes of specific size in the
photosphere~\citep{bunteInterchangeInstabilitySolar1993} and in the buoyancy of
flux tubes~\citep{schuesslerInterchangeInstabilitySmall1984}. However, the
flute instability specifically is more commonly studied in fusion
contexts~\citep{mikhailovskiiInstabilitiesConfinedPlasma1998,zhengAdvancedTokamakStability2015,wessonHydromagneticStabilityTokamaks1978}.
In fusion, the focus is generally on understanding how a particular plasma
device may be stabilised to the instability in particular geometries such as
that of the mirror machine~\citep{jungwirthTheoryFluteInstability1965} or in
toroidal geometries such as the
tokomak~\citep{shafranovFluteInstabilityCurrentcarrying1968}. The resistive
flute instability (also known as the resistive interchange instability) can be
excited even when the ideal flute instability is stabilised. As a result, this
has been given significantly more
attention~\citet{johnsonResistiveInterchangesNegativeV1967,correa-restrepoResistiveBallooningModes1983}.
While this body of research is useful and applicable in solar contexts, it is
mostly limited to the study of the stability and linear development of the
flute instability, the nonlinear development being of secondary importance in
the investigation of fusion devices. More detailed investigations of its
nonlinear development is required to understand its importance in the context
of coronal dynamics and coronal heating. The development of the flute
instability and its interaction with the simultaneously growing kink
instability is the main focus of this work and the experiments described here
represent an initial exploration into the nonlinear flute instability in the
solar corona.

In addition to our main goal, of particular interest here is the effect of
anisotropic plasma viscosity, which in the following is found to strongly
influence the growth of the flute instability. It is well known that viscosity
in magnetised plasmas (such as those which make up the solar corona) is
anisotropic and strongly dependent on the strength and direction of the local
magnetic
field~\citep{hollwegViscosityChewGoldbergerLowEquations1986,hollwegViscosityMagnetizedPlasma1985,braginskiiTransportProcessesPlasma1965}.
To take this into account, MacTaggaret et al. developed a phenomenological
model of anisotropic
viscosity~\citep{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017} that
captures the main physics of viscosity in the solar corona as outlined in the
analysis of~\cite{braginskiiTransportProcessesPlasma1965}, namely parallel
viscosity in regions of strong field  strength and isotropic viscosity in
regions of very weak or zero field strength. For brevity, we will refer to this
model of viscosity as ``the switching model''.
In~\citet{quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}
we implemented the switching model as a module for the widely-used general MHD
code Lare3d~\citep{arberStaggeredGridLagrangian2001}, and demonstrated
significant effects of anisotropic viscosity on the development of the
nonlinear MHD kink instability and the Kelvin-Helmholtz instability. More
generally, the interest in anisotropic viscosity stems from the open question
of which heating mechanism (viscous or Ohmic) is dominant in the solar
corona~\citep{klimchukSolvingCoronalHeating2006}, an important facet of solving
the coronal heating problem. Using scaling laws, it has been suggested that
viscous heating (generated through anisotropic viscosity) can dwarf that of
Ohmic
heating~\citep{craigAnisotropicViscousDissipation2009a,litvinenkoViscousEnergyDissipation2005}.
However, due to computational and observational limitations, this cannot be
directly tested, and so the influence of other factors such as small scale
instabilities and turbulence is relatively
unknown~\citet{klimchukSolvingCoronalHeating2006}. In addition to directly
heating the plasma, viscosity plays a part in the damping of instabilities and
waves~\citep{rudermanSlowSurfaceWave2000}. It is this effect we are most
interested in here, and it shall be reported that the use of anisotropic
viscosity permits the growth of the flute instability, which is otherwise
strongly damped by isotropic viscosity.

\rev{It will be shown in our results that the flute instability is driven by
a pressure gradient generated through Ohmic heating, so the value of
resistivity is an important parameter in the development of the instability. In
addition to exploring the effect of viscosity model on the flute instability,
we also explore the effect of varying resistivity. Ideally, our simulations
would be performed using a realistic coronal resistivity of approximately
$10^{-8}$~\citep{craigAnisotropicViscousDissipation2009a} however due to the
dissipative nature of numerical schemes (particularly when using shock
capturing techniques) this is computationally infeasible. In order to provide
some sense of the scaling expected with resistivity, to allow us to speculate
about the expected results at coronal values, we perform our simulations at
two, computationally accessible values of resistivity, $10^{-3}$ and $10^{-4}$.
This is not intended to replace a full parameter study so we will speculate
whether our results generalise to lower values of resistivity but will not
attempt to quantify any scaling laws or suggest concrete trends.}

This paper is organised as follows. Section~\ref{sec-flute-intro} introduces
the flute instability and relevant linear analyses,
section~\ref{sec-numerical-setup} describes the governing equations, coronal
loop model and details of the numerical parameters, section~\ref{sec-results}
presents the overall development of the flute instability, before discussing
features unique to the two values of resistivity studied,
section~\ref{sec-discussion} discusses the limitations of the simulations, with
suggestions for future work, and section~\ref{sec-conclusions} presents our
conclusions in the wider context of coronal heating.

\subsection{The flute instability}
\label{sec-flute-intro}

In general, the stability of a cylindrical twisted magnetic flux tube is analysed using perturbations of the form
\begin{equation}
  \label{eq:kink_perturbation}
\xi(r, \theta, z) = \xi(r) e^{i(m\theta + kz)},
\end{equation}
where $m$ and $k$ are the wavenumbers in the azimuthal and axial directions,
$\theta$ and $z$, respectively and $r$ is the radial coordinate in cylindrical
polars. The helical kink instability occurs for perturbations where $m=1,
k\ne0$ and is the only instability of this form which is a body instability,
that is it moves the entire body of the flux tube. Perturbations where $m>1$
are termed flute or interchange instabilities.

@@

The stability of an infinite cylindrical flux tube to perturbations of the
form~\eqref{eq:kink_perturbation} is given by the classical Suydam's
criterion~\citep{suydamStabilityLinearPinch1958}
\begin{equation}
  \label{eq:suydams_criterion}
\frac{B_z^2 S^2}{4} + 2 r p' > 0,
\end{equation}
where $S = r q'/q$ is a measure of the shear, $q = 2\pi r B_z / L B_{\theta}$
is the safety factor for a flux tube of length $L$ and a prime denotes
differentiation with respect to
$r$~\citep{mikhailovskiiInstabilitiesConfinedPlasma1998}. This applies to both
flute and kink instabilities although many additional effects such as
line-tying are not incorporated into the corresponding linear analysis. The
effect of line-tying on the kink instability is investigated
in~\citet{hoodKinkInstabilitySolar1979}. Where~\eqref{eq:suydams_criterion} is
not satisfied, the flux tube may be unstable to perturbations of the
form~\eqref{eq:kink_perturbation}. When $m>1$, the perturbations remain local
to resonant surfaces given by~\eqref{eq:resonant_surface}. When Suydam's
criterion is satisfied and the flux tube is linearly stable, it may still be
unstable to non-local perturbations, where the shear and pressure are small
enough that interchange perturbations do not need to
satisfy~\eqref{eq:resonant_surface}. Additionally, the inclusion of resistivity
generally reduces the stabilising effect of the shear, permitting growth of
a resistive interchange mode, albeit at a slower rate than that of the ideal
instability~\citep{mikhailovskiiInstabilitiesConfinedPlasma1998}. It will be
found that the ideal linear analysis
of~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998} is sufficient for
understanding the flute instabilities investigated here since the associated
flux tubes adequately fail the criterion~\eqref{eq:suydams_criterion}.

While Suydam's condition gives an indication of the stability of a flux tube to
a given perturbation, the linear growth rate of the ideal flute instability
$\gamma$ can be found via a stability analysis analogous to that of the
Rayleigh-Taylor instability
(see~\citep[][]{goldstonIntroductionPlasmaPhysics2020}) and is given by
\begin{equation}
  \label{eq:fluting_growth_rate}
\gamma^2 = \frac{2|\nabla p|}{\rho R_c},
\end{equation}
where $R_c$ is the radius of curvature of the magnetic field. This equation
only applies when the pressure gradient and radius of curvature vector are in
the same direction, that is the plasma is constrained by a concave magnetic
field such that the pressure forces and magnetic tension forces are in
competition. In a cylindrical, twisted flux tube, the field is always concave
towards the central axis of the tube, so any inwardly directed pressure
gradient is potentially unstable to fluting.

@@

The stability criterion \eqref{eq:suydams_criterion} and the linear growth rate
approximation \eqref{eq:fluting_growth_rate2} are useful only as a guide and
for approximate analysis of the numerical simulations presented in this work.
The precise form of the equilibrium state and the perturbations needed for the
validity of \eqref{eq:suydams_criterion} and \eqref{eq:fluting_growth_rate2}
were used in~\citet{quinnEffectAnisotropicViscosity2020}. In contrast, in the
experiments reported in the following the system is driven and instabilities
occur spontaneously due to random perturbation. As a result of the driving, the
flux tube is also not in static equilibrium initially.

@@

Expression \eqref{eq:switching_model} is identical to the strong field
approximation of the general anisotropic viscosity tensor derived
in~\citet{braginskiiTransportProcessesPlasma1965}. Expressions
\eqref{eq:isotropic_viscous_tensor} and \eqref{eq:switching_model} arise as
asymptotic limits of the more general switching model used in our earlier works~\citep{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017,quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}
which, includes both isotropic and anisotropic contributions and can switch
gradually between them depending on the strength of the magnetic field at
a given spacio-temporal location. For example, in the vicinity of a null point
where the magnetic field becomes weak the isotropic viscosity contribution
becomes dominant in the switching model. Switching between the two limit cases
is not relevant in the present study where the variations in the magnetic field
are not significantly large.

@@

The driver velocity is set to $u_0 = 0.15$, the normalising factor is $u_{r0}
= 2.08$, and setting $r_d = 5$ corresponds to a driver constrained to $r<1$ and
with a peak velocity at $r\approx 0.38$. The ramping time is set to $t_r = 2$,
resulting in an acceleration from $0$ to $u_0$ over approximately $5$ Alfv\'en
times. These driver parameters correspond to a peak rotational period of $T_R
= 15.92$, the length of time taken for one full turn to be injected by a single
driver. Both drivers result in twist being added at a rate of $2\pi$ every
$7.96$ Alfv\'en times. The twist profile across the entire flux tube develops
in such a way that by $t\approx 20$, it is qualitatively similar to those
studied
in~\citet{quinnEffectAnisotropicViscosity2020,hoodCoronalHeatingMagnetic2009,barefordShockHeatingNumerical2015}
however the length of the flux tubes differs significantly. This configuration
produces a $z$-directed tube of increasingly twisted magnetic field that
eventually becomes unstable to both the flute instability and the helical kink
instability.

The problem formulated above is solved numerically using the staggered-grid,
Lagrangian–Eulerian remap code for 3D MHD simulations Lare3D
of~\cite{arberStaggeredGridLagrangian2001} where a  new module for anisotropic
viscosity has been included as detailed
in~\citet{quinnKelvinHelmholtzInstabilityCollapse2021}. The resolution used in
the current work is  $512$ grid points per dimension, comparable to the highest
resolution kink instability studies of~\cite{hoodCoronalHeatingMagnetic2009} or
medium resolution studies of~\cite{barefordShockHeatingNumerical2015}.

@@

We focus the attention on two selected pairs of  simulations, one pair where
the background resistivity is set to $\eta=10^{-3}$ and another where
$\eta=10^{-4}$. As in~\citet{quinnEffectAnisotropicViscosity2020}, only
background resistivity is used. Each pair consists of one simulation using
isotropic viscosity~\eqref{eq:isotropic_viscous_tensor} and another one using
the anisotropic model~\eqref{eq:switching_model}. The value of viscosity is set
to $\nu = 10^{-4}$ in all cases.

@@

As the field continues to be twisted, currents form, due to the local shear in
the magnetic field, and heat the plasma through Ohmic dissipation. Due to the
radial form of the driver, the magnitude of the current density is greatest at
the axis of the tube, then decreases radially outwards as seen in
figure~\ref{fig:current_profiles}. The orientation of the twisting produces
a current flowing in the negative $z$-direction for $r\lessapprox0.5$. At $r
\approx 0.5$ (corresponding to the radius of peak driving velocity) the current
switches orientation and is in the positive $z$-direction in a shell where
$0.5\lessapprox r \lessapprox 0.8$. This form of a twisted field with an inner
core of current in one direction surrounded by a shell of oppositely-directed
current is similar to the current configuration arising due to the field
prescribed in~\citet{quinnEffectAnisotropicViscosity2020}.

@@

The faster growth of the kink compared to that measured in~\citet{quinnEffectAnisotropicViscosity2020} is attributed to the relative
aspect ratios of the flux tubes. The tube prescribed in~\citet{quinnEffectAnisotropicViscosity2020} has an aspect ratio of
approximately $20$ compared to the tube studied here which has an aspect ratio
of approximately $4$. While the total twist is similar in both tubes (after the
drivers have injected twist up to $t\approx20$) the small aspect ratio results
in more turns per unit length, leading to a faster growing instability.

>>>>

%% --- user packages ---
\usepackage{csquotes}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{grffile}
\usepackage{overpic}
\usepackage{float}
\usepackage{enumitem}
\usepackage{amsfonts}
\usepackage{easy-todo}
\usepackage{listings}
%\usepackage[citecolor=blue]{hyperref}
\usepackage{lineno}
\usepackage[normalem]{ulem}

@@

\newcommand{\rev}[1]{{\color{red} {#1}}}
\newcommand{\newold}[2]{{#1}{}}
\newcommand{\revcite}[1]{{\color{red} \underline{#1}}}

@@

\begin{abstract}
Magnetic flux tubes such as those in the solar corona are subject to a number
of instabilities. Important among them is the kink instability which plays
a central part in the nanoflare theory of coronal heating, and for this reason
in numerical simulations it is usually induced by tightly-controlled
perturbations and studied in isolation. In contrast, we find that
fluting modes of instability are readily excited when disturbances are
introduced in our magnetohydrodynamic flux tube simulations by dynamic
twisting of the flow at the boundaries. We also find that the flute instability, which has been 
theorised but rarely observed in the coronal context,  is strongly enhanced
when plasma viscosity is assumed anisotropic. We proceed to investigate the
co-existence and competition between flute and kink instabilities for a range
of values of the resistivity and of the parameters of the anisotropic and
isotropic models of viscosity. We conclude that while the flute instability
cannot prevent the kink from ultimately dominating, it can significantly delay
its development especially at strong viscous anisotropy induced by intense
magnetic fields.
\end{abstract}

@@

The helical kink instability is a form of ideal magnetohydrodynamic (MHD)
instability which occurs in highly twisted magnetic flux tubes such as those
making up much of the solar corona~\citep{realeCoronalLoopsObservations2014}
and has been well studied in the coronal
context~\citep{hoodKinkInstabilitySolar1979,hoodCoronalHeatingMagnetic2009,browningSolarCoronalHeating2003b,barefordShockHeatingNumerical2015,quinnEffectAnisotropicViscosity2020}.
Given its energetic nonlinear development, it is considered a potential
mechanism for heating the solar corona through the theory of
nanoflares~\citep{klimchukSolvingCoronalHeating2006,browningMechanismsSolarCoronal1991}
and a key mechanism in the production of solar
flares~\citep{hoodKinkInstabilitySolar1979}. \rev{Our previous work}
investigated a twisted magnetic flux tube already linearly unstable to
the 
helical kink instability, focussing specifically on the effect of anisotropic
viscosity on the nonlinear
dynamics~\citep{quinnEffectAnisotropicViscosity2020}.
\rev{There and in} most other investigations of
the kink instability e.g.~\revcite{\citep{hoodCoronalHeatingMagnetic2009}},
a perturbation is applied to an already significantly twisted flux tube. An
alternative way to excite the instability (and the way employed here) is to
start with an initially straight field and apply twisting motions at the
boundaries to form a twisted flux tube which eventually becomes unstable. This
kind of dynamic excitation of the kink instability \rev{}
%is useful in that it
represents more closely the actual evolution of magnetic flux tubes and the
associated instabilities in the solar corona. In our simulations, the dynamic
twisting of the flux tube reveals an additional instability, the flute
instability, which has been theorised, for example
in~\revcite{\citep{priestMagnetohydrodynamicsSun2013}}. While oscillations resembling
flute perturbations have been found in simulations of coronal
loops~\citep{terradasEffectMagneticTwist2018}, to our knowledge, this is the
first time the flute instability has been investigated computationally in
a coronal context.

@@

In other solar contexts, interchange instabilities can be found in the form of
ballooning modes in arcades~\citep{hoodBallooningInstabilitiesSolar1986}, as
the instability which forms tubes of specific size in the
photosphere~\citep{bunteInterchangeInstabilitySolar1993}, and in the buoyancy of
flux tubes~\citep{schuesslerInterchangeInstabilitySmall1984}. However, the
flute instability specifically is more commonly studied in fusion
contexts~\citep{mikhailovskiiInstabilitiesConfinedPlasma1998,zhengAdvancedTokamakStability2015,wessonHydromagneticStabilityTokamaks1978}.
In fusion, the focus is generally on understanding how a particular plasma
device may be stabilised to the instability in particular geometries such as
that of the mirror machine~\citep{jungwirthTheoryFluteInstability1965} or in
toroidal geometries such as the
tokamak~\citep{shafranovFluteInstabilityCurrentcarrying1968}. The resistive
flute instability (also known as the resistive interchange instability) can be
excited even when the ideal flute instability is stabilised. As a result, this
has been given significantly more
attention~\revcite{\citep{johnsonResistiveInterchangesNegativeV1967,correa-restrepoResistiveBallooningModes1983}}.
While this body of research is useful and applicable in solar contexts, it is
mostly limited to the study of the stability and linear development of the
flute instability, the nonlinear development being of secondary importance in
the investigation of fusion devices. More detailed investigations of its
nonlinear development is required to understand its importance in the context
of coronal dynamics and coronal heating. The development of the flute
instability and its interaction with the simultaneously growing kink
instability is the main focus of this work and the experiments described here
represent an initial exploration into the nonlinear flute instability in the
solar corona.

In addition to our main goal, of particular interest here is the effect of
anisotropic plasma viscosity, which in the following is found to strongly
influence the growth of the flute instability. It is well known that viscosity
in magnetised plasmas (such as those which make up the solar corona) is
anisotropic and strongly dependent on the strength and direction of the local
magnetic
field~\citep{hollwegViscosityChewGoldbergerLowEquations1986,hollwegViscosityMagnetizedPlasma1985,braginskiiTransportProcessesPlasma1965}.
To take this into account,
\revcite{\citet{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017}}
developed a phenomenological model of anisotropic  viscosity
%\rev{\citep{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017}}
that captures the main physics of viscosity in the solar corona as outlined in the
analysis of~\cite{braginskiiTransportProcessesPlasma1965}, namely parallel
viscosity in regions of strong field  strength and isotropic viscosity in
regions of very weak or zero field strength. For brevity, we will refer to this
model of viscosity as ``the switching model''.
In \revcite{\citep{quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}}
we implemented the switching model as a module for the widely-used general MHD
code Lare3d~\citep{arberStaggeredGridLagrangian2001}, and demonstrated
significant effects of anisotropic viscosity on the development of the
nonlinear MHD kink instability and the Kelvin-Helmholtz instability. More
generally, the interest in anisotropic viscosity stems from the open question
of which heating mechanism (viscous or Ohmic) is dominant in the solar
corona~\citep{klimchukSolvingCoronalHeating2006}, an important facet of solving
the coronal heating problem. Using scaling laws, it has been suggested that
viscous heating (generated through anisotropic viscosity) can dwarf that of
Ohmic
heating~\citep{craigAnisotropicViscousDissipation2009a,litvinenkoViscousEnergyDissipation2005}.
However, due to computational and observational limitations, this cannot be
directly tested, and so the influence of other factors such as small scale
instabilities and turbulence is relatively
unknown~\revcite{\citep{klimchukSolvingCoronalHeating2006}}. In addition to directly
heating the plasma, viscosity plays a part in the damping of instabilities and
waves~\citep{rudermanSlowSurfaceWave2000}. It is this effect we are most
interested in here, and it shall be reported that the use of anisotropic
viscosity permits the growth of the flute instability, which is otherwise
strongly damped by isotropic viscosity.

\rev{The value of plasma resistivity strongly affects the development of
the flute instability because the instability is due to pressure
gradients generated through Ohmic heating as will be shown further below.
Ideally, our simulations would be performed using a realistic coronal
resistivity values of approximately
$10^{-8}$~\citep{craigAnisotropicViscousDissipation2009a}, however, due
to the dissipative nature of numerical schemes (particularly when using shock
capturing techniques) this is computationally infeasible.
To overcome this limitation and to attempt to extrapolate results
towards more realistic values, we perform and compare simulations at 
two, computationally accessible, resistivity values, $10^{-3}$ and
$10^{-4}$. This comparison runs as an additional theme of the paper,
even if not quite a primary aim.}
% To overcome this limitation, we perform and compare simulations at
% two, computationally accessible, values of resistivity, $10^{-3}$ and
% $10^{-4}$, and this comparison runs as an additional theme of the paper.
% While this does not constitute a detailed parameter study of
% resistivity (which is expensive) it allows us to speculate as to how
% results might extrapolate to more realistic values.}
%
% \rev{It will be shown in our results that the flute instability is driven by
% a pressure gradient generated through Ohmic heating, so the value of
% resistivity is an important parameter in the development of the instability. In
% addition to exploring the effect of viscosity model on the flute instability,
% we also explore the effect of varying resistivity. Ideally, our simulations
% would be performed using a realistic coronal resistivity of approximately
% $10^{-8}$~\citep{craigAnisotropicViscousDissipation2009a} however due to the
% dissipative nature of numerical schemes (particularly when using shock
% capturing techniques) this is computationally infeasible. In order to provide
% some sense of the scaling expected with resistivity, to allow us to speculate
% about the expected results at coronal values, we perform our simulations at
% two, computationally accessible values of resistivity, $10^{-3}$ and $10^{-4}$.
% This is not intended to replace a full parameter study so we will speculate
% whether our results generalise to lower values of resistivity but will not
% attempt to quantify any scaling laws or suggest concrete trends.}

Our article is organised as follows. Section~\ref{sec-flute-intro} introduces
the flute instability and \rev{recalls relevant background on its linear stability analysis}.
Section~\ref{sec-numerical-setup} describes the governing equations,
the coronal loop model and the numerical parameters used. Section~\ref{sec-results}
presents the overall development of the flute instability before
\rev{proceeding to compare simulations in the cases of various viscous
anisotropy and various resistivity values. Here results are
organised by resistivity values as this allows to contrast isotropic
and anisotropic viscosity cases more directly.}
Section~\ref{sec-discussion} discusses the limitations of the simulations, with
suggestions for future work, and section~\ref{sec-conclusions} presents our
conclusions in the wider context of coronal heating.

\section{The flute instability}
\label{sec-flute-intro}

In general, the stability of a cylindrical twisted magnetic flux tube is analysed using perturbations of the form
\begin{equation}
  \label{eq:kink_perturbation}
\xi(r, \theta, z) = \xi(r) e^{i(m\theta + kz)},
\end{equation}
where $m$ and $k$ are the wavenumbers in the azimuthal and axial directions,
$\theta$ and $z$, respectively and $r$ is the radial coordinate in cylindrical
polars. The helical kink instability occurs for perturbations where $m=1,
k\ne0$ and is the only instability of this form which is a body instability,
i.e.{} it moves the entire body of the flux tube. Perturbations where $m>1$
are termed flute or interchange instabilities.

@@

The stability of an infinite cylindrical flux tube to perturbations of the
form~\eqref{eq:kink_perturbation} is given by the classical Suydam's
criterion~\citep{suydamStabilityLinearPinch1958}
\begin{equation}
  \label{eq:suydams_criterion}
\frac{B_z^2 S^2}{4} + 2 r p' > 0,
\end{equation}
where $S = r q'/q$ is a measure of the shear, $q = 2\pi r B_z / L B_{\theta}$
is the safety factor for a flux tube of length $L$ and a prime denotes
differentiation with respect to
$r$~\citep{mikhailovskiiInstabilitiesConfinedPlasma1998}. This applies to both
flute and kink instabilities although many additional effects such as
line-tying are not incorporated into the corresponding linear analysis. The
effect of line-tying on the kink instability is investigated
in \revcite{\citep{hoodKinkInstabilitySolar1979}}. Where~\eqref{eq:suydams_criterion} is
not satisfied, the flux tube may be unstable to perturbations of the
form~\eqref{eq:kink_perturbation}. When $m>1$, the perturbations remain local
to resonant surfaces given by~\eqref{eq:resonant_surface}. When Suydam's
criterion is satisfied and the flux tube is linearly stable, it may still be
unstable to non-local perturbations, where the shear and pressure are small
enough that interchange perturbations do not need to
satisfy~\eqref{eq:resonant_surface}. Additionally, the inclusion of resistivity
generally reduces the stabilising effect of the shear, permitting growth of
a resistive interchange mode, albeit at a slower rate than that of the ideal
instability~\citep{mikhailovskiiInstabilitiesConfinedPlasma1998}. It will be
found that the ideal linear analysis
of~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998} is sufficient for
understanding the flute instabilities investigated here since the associated
flux tubes adequately fail the criterion~\eqref{eq:suydams_criterion}.

While Suydam's condition gives an indication of the stability of a flux tube to
a given perturbation, the linear growth rate of the ideal flute instability
$\gamma$ can be found via a stability analysis analogous to that of the
Rayleigh-Taylor instability
see~\citep[][]{goldstonIntroductionPlasmaPhysics2020}, and is given by
\begin{equation}
  \label{eq:fluting_growth_rate}
\gamma^2 = \frac{2|\nabla p|}{\rho R_c},
\end{equation}
where $R_c$ is the radius of curvature of the magnetic field. This equation
only applies when the pressure gradient and radius of curvature vector are in
the same direction, that is the plasma is constrained by a concave magnetic
field such that the pressure forces and magnetic tension forces are in
competition. In a cylindrical, twisted flux tube, the field is always concave
towards the central axis of the tube, so any inwardly directed pressure
gradient is potentially unstable to fluting.

@@

The stability criterion \eqref{eq:suydams_criterion} and the linear growth rate
approximation \eqref{eq:fluting_growth_rate2} are useful only as a guide and
for approximate analysis of the numerical simulations presented in this work.
The precise form of the equilibrium state and the perturbations needed for the
validity of \eqref{eq:suydams_criterion} and \eqref{eq:fluting_growth_rate2}
were used in~\revcite{\citep{quinnEffectAnisotropicViscosity2020}}. In contrast, in the
experiments reported in the following the system is driven and instabilities
occur spontaneously due to random perturbations. As a result of the driving, the
flux tube is also not in static equilibrium initially.

@@

Expression \eqref{eq:switching_model} is identical to the strong field
approximation of the general anisotropic viscosity tensor derived
by \revcite{\citet{braginskiiTransportProcessesPlasma1965}}. Expressions
\eqref{eq:isotropic_viscous_tensor} and \eqref{eq:switching_model} arise as
asymptotic limits of the more general switching model used in our earlier works~\citep{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017,quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}
which, includes both isotropic and anisotropic contributions and can switch
gradually between them depending on the strength of the magnetic field at
a given spacio-temporal location. For example, in the vicinity of a null point
where the magnetic field becomes weak the isotropic viscosity contribution
becomes dominant in the switching model. Switching between the two limit cases
is not relevant in the present study where the variations in the magnetic field
are not significantly large.

@@

The driver velocity is set to $u_0 = 0.15$, the normalising factor is $u_{r0}
= 2.08$, and setting $r_d = 5$ corresponds to a driver constrained to $r<1$ and
with a peak velocity at $r\approx 0.38$. The ramping time is set to $t_r = 2$,
resulting in an acceleration from $0$ to $u_0$ over approximately $5$ Alfv\'en
times. These driver parameters correspond to a peak rotational period of $T_R
= 15.92$, the length of time taken for one full turn to be injected by a single
driver. Both drivers result in twist being added at a rate of $2\pi$ every
$7.96$ Alfv\'en times. The twist profile across the entire flux tube develops
in such a way that by $t\approx 20$, it is qualitatively similar to those
studied
in~\citep{quinnEffectAnisotropicViscosity2020,hoodCoronalHeatingMagnetic2009,barefordShockHeatingNumerical2015}
however the length of the flux tubes differs significantly. This configuration
produces a $z$-directed tube of increasingly twisted magnetic field that
eventually becomes unstable to both the flute instability and the helical kink
instability.

The problem formulated above is solved numerically using the staggered-grid,
Lagrangian–Eulerian remap code for 3D MHD simulations Lare3D
of~\cite{arberStaggeredGridLagrangian2001} where a  new module for anisotropic
viscosity has been included as detailed
in~\citep{quinnKelvinHelmholtzInstabilityCollapse2021}. The resolution used in
the current work is  $512$ grid points per dimension, comparable to the highest
resolution kink instability studies of~\cite{hoodCoronalHeatingMagnetic2009} or
medium resolution studies of~\cite{barefordShockHeatingNumerical2015}.

@@

We focus the attention on two selected pairs of  simulations, one pair where
the background resistivity is set to $\eta=10^{-3}$ and another where
$\eta=10^{-4}$. As in~\citep{quinnEffectAnisotropicViscosity2020}, only
background resistivity is used. Each pair consists of one simulation using
isotropic viscosity~\eqref{eq:isotropic_viscous_tensor} and another one using
the anisotropic model~\eqref{eq:switching_model}. The value of viscosity is set
to $\nu = 10^{-4}$ in all cases.

@@

As the field continues to be twisted, currents form, due to the local shear in
the magnetic field, and heat the plasma through Ohmic dissipation. Due to the
radial form of the driver, the magnitude of the current density is greatest at
the axis of the tube, then decreases radially outwards as seen in
figure~\ref{fig:current_profiles}. The orientation of the twisting produces
a current flowing in the negative $z$-direction for $r\lessapprox0.5$. At $r
\approx 0.5$ (corresponding to the radius of peak driving velocity) the current
switches orientation and is in the positive $z$-direction in a shell where
$0.5\lessapprox r \lessapprox 0.8$. This form of a twisted field with an inner
core of current in one direction surrounded by a shell of oppositely-directed
current is similar to the current configuration arising due to the field
prescribed in~\citep{quinnEffectAnisotropicViscosity2020}.

@@

The faster growth of the kink compared to that measured in~\citep{quinnEffectAnisotropicViscosity2020} is attributed to the relative
aspect ratios of the flux tubes. The tube prescribed in~\citep{quinnEffectAnisotropicViscosity2020} has an aspect ratio of
approximately $20$ compared to the tube studied here which has an aspect ratio
of approximately $4$. While the total twist is similar in both tubes (after the
drivers have injected twist up to $t\approx20$) the small aspect ratio results
in more turns per unit length, leading to a faster growing instability.
